A Software Engineering Perspective on \\ OpenMP and PThreads

### 2017-09-08 rebuttal/rebuttal.md

## Paper Reviews and Rebuttal

### Review Summary

| Paper Merit   | Reviewer Expertise |
| :------------ | :----------------- |
| Accept        | Knowledgeable      |
| Weak Reject   | Knowledgeable      |
| Strong Accept | Some Familiarity   |

### Reviewer Questions and Rebuttal

### Reviewer #9A

The following explicit questions need to be addressed:

1. What were the learning objectives of the course and research project?
2. Did you have ethical approval to do a study like this? You mention consent
   but not much more than that
3. You mention significant, but you don't have evidence to support this claim.
   You need to show statistical analysis for your data.
4. The graphs are hard to read. Consider using a different colour palette.
5. Need to add limitation or threats to validity for your research approach.
6. Do you have any comments on generalizing this kind of research approach to
   study other aspects of programming?

### Reviewer #9B

The following explicit questions need to be addressed:

#### Explicit Questions

##### Major Issues

1. The description of the programming task is somewhat vague and it is unclear
   if it went through any validation for appropriateness to measure students'
   ability. What was the exact task (the authors mention the Mandelbrot Set
   they provided as an example, but do not expand on their "simple problem").
   There is no way for the reader to determine how appropriate this task was
   without knowing the problem, so perhaps listing the question/problem itself
   would be best.
2. The reviewer assumes that the programming assignment and questionnaires were
   solo activities. However, this was not explicitly described. (Since the
   authors described this course as a place where undergraduate students and
   graduate students could share their knowledge).
3. How were the questions developed, and what measures did the authors take to
   make sure they were appropriate to measure what they wanted to measure?
4. Were the questions listed in Portuguese (the reviewer believes this might be
   an option as this study took place in Brazil) or English?
5. Some of the questions in Table 1 are vague and the students might interpret
   them in different ways. For example, what level of commitment/experience
   does "Had contact with the APIs before?" assume? What does "the API" mean?
   The same question goes for what "learned" means in the following two
   questions.
6. In Table 3, what is the difference between questions 2 and 3? "Was simpler
   to use" vs. "Was easier to use".
7. The reviewer does not believe that the major finding of this paper is
   accurately described and cannot agree with the author's statement in the
   conclusion that 52.2% of students performing better using Pthreads
   constitutes a "majority."

##### Minor Issues

1. Some of the statements in the introduction and the descriptions of OpenMP
   and Pthreads can use additional references (e.g., "slowdowns on Moore's law
   and the end of Dennard scaling").
2. Recommend changing "Listing X" to "Figure X".
3. Recommend displaying "Listing 1" in one column instead of splitting it in
   the middle.
4. Recommend using active voice instead of passive voice (e.g., "Different
   approaches have been proposed to study..."
5. Recommend not using colors for Figures since some readers might not be able
   to distinguish them (or someone might print it out in black-and-white).
6. Table 1. The second question has a typo "Had contat [sic] with the API///"

#### Other Comments

- Comment:
> The authors did answer their three research questions. However, the authors
> over-claim one of their major results (see #7, below), along with saying some
> of their results are "significant" without providing statistical measure.

### Reviewer #9C

#### Explicit Questions

1. Why do you call it a Likert-like scale?
2. How were the APIs introduced in the course?
3. What were the exercises?
4. Is it possible that differences in recognition depend on the presentation of
   the APIs or teaching material, etc.?

#### Other Comments

- Comment
> I personally think that a much larger discussion section is required that
> actually discusses the chosen method and the conclusions drawn from it.

- Comment:
> Since you have a Likert-Scale, you can apply non-parametric tests on it in
> order to check, whether group A or B is different. Descriptive data and raw
> measurement is fine, because it gives readers a change to get an idea about
> the data, but only relying on such things is not enough. Having said this, I
> do not think that you can conclude something from the present paper, because
> the conclusion is just based on "looking at descriptive data".

- Comment:
> Please fix the statistical part of the paper. Nothing would be more annoying
> than having a paper published whose interpretation depends on results that
> would be different in case appropriate analysis techniques would have been
> applied.

### 2017-09-08 rebuttal/rebuttal.md

##### Minor Issues

1. Some of the statements in the introduction and the descriptions of OpenMP
   and Pthreads can use additional references (e.g., "slowdowns on Moore's law
   and the end of Dennard scaling").
2. Recommend changing "Listing X" to "Figure X".
3. Recommend displaying "Listing 1" in one column instead of splitting it in
   the middle.
4. Recommend using active voice instead of passive voice (e.g., "Different
   approaches have been proposed to study..."
5. Recommend not using colors for Figures since some readers might not be able
   to distinguish them (or someone might print it out in black-and-white).
6. Table 1. The second question has a typo "Had contat [sic] with the API///"

#### Other Comments

1. The authors did answer their three research questions. However, the authors
   over-claim one of their major results (see #7, below), along with saying some
   of their results are "significant" without providing statistical measure.

@@

#### Explicit Questions

1. Why do you call it a Likert-like scale?
2. How were the APIs introduced in the course?
3. What were the exercises?
4. Is it possible that differences in recognition depend on the presentation of
   the APIs or teaching material, etc.?

#### Other Comments

1. I personally think that a much larger discussion section is required that
  actually discusses the chosen method and the conclusions drawn from it.
2. Since you have a Likert-Scale, you can apply non-parametric tests on it in
   order to check, whether group A or B is different. Descriptive data and raw
   measurement is fine, because it gives readers a change to get an idea about
   the data, but only relying on such things is not enough. Having said this, I
   do not think that you can conclude something from the present paper, because
   the conclusion is just based on "looking at descriptive data".
3. Please fix the statistical part of the paper. Nothing would be more annoying
   than having a paper published whose interpretation depends on results that
   would be different in case appropriate analysis techniques would have been
   applied.

### 2017-09-21 rebuttal/rebuttal.md

### Reviewer Questions and Rebuttal

### Reviewer #9A

#### Explicit Questions

1. What were the learning objectives of the course and research project? (TODO: Alfredo)
2. Did you have ethical approval to do a study like this? You mention consent
   but not much more than that (TODO: Paulo)
3. You mention significant, but you don't have evidence to support this claim.
   You need to show statistical analysis for your data (TODO: Pedro)
4. The graphs are hard to read. Consider using a different colour palette (TODO: Pedro)
5. Need to add limitation or threats to validity for your research approach (TODO: Raphael)
6. Do you have any comments on generalizing this kind of research approach to
   study other aspects of programming? (TODO: Alfredo)

@@

### Reviewer #9B

#### Explicit Questions

##### Major Issues

1. The description of the programming task is somewhat vague and it is unclear
   if it went through any validation for appropriateness to measure students'
   ability. What was the exact task (the authors mention the Mandelbrot Set
   they provided as an example, but do not expand on their "simple problem").
   There is no way for the reader to determine how appropriate this task was
   without knowing the problem, so perhaps listing the question/problem itself
   would be best. (TODO: Pedro)
2. The reviewer assumes that the programming assignment and questionnaires were
   solo activities. However, this was not explicitly described. (Since the
   authors described this course as a place where undergraduate students and
   graduate students could share their knowledge). (TODO: Alfredo)
3. How were the questions developed, and what measures did the authors take to
   make sure they were appropriate to measure what they wanted to measure? (DONE: ele pode ter razão, não temos o que fazer)
4. Were the questions listed in Portuguese (the reviewer believes this might be
   an option as this study took place in Brazil) or English? (TODO: Pedro)
5. Some of the questions in Table 1 are vague and the students might interpret
   them in different ways. For example, what level of commitment/experience
   does "Had contact with the APIs before?" assume? What does "the API" mean?
   The same question goes for what "learned" means in the following two
   questions.(DONE: ele pode ter razão, não temos o que fazer)
6. In Table 3, what is the difference between questions 2 and 3? "Was simpler
   to use" vs. "Was easier to use". (TODO: Pedro)
7. The reviewer does not believe that the major finding of this paper is
   accurately described and cannot agree with the author's statement in the
   conclusion that 52.2% of students performing better using Pthreads
   constitutes a "majority." (TODO: Pedro, mas acho que se pensarmos juntos mitigaremos)

@@

##### Minor Issues

1. Some of the statements in the introduction and the descriptions of OpenMP
   and Pthreads can use additional references (e.g., "slowdowns on Moore's law
   and the end of Dennard scaling"). (TODO: Raphael)
2. Recommend changing "Listing X" to "Figure X". (TODO: Paulo)
3. Recommend displaying "Listing 1" in one column instead of splitting it in
   the middle. (TODO: Paulo)
4. Recommend using active voice instead of passive voice (e.g., "Different
   approaches have been proposed to study..." (TODO: Paulo)
5. Recommend not using colors for Figures since some readers might not be able
   to distinguish them (or someone might print it out in black-and-white). (TODO: Pedro, relacionado com item 4 do revisão 1)
6. Table 1. The second question has a typo "Had contat [sic] with the API///" (TODO: Paulo)

@@

1. The authors did answer their three research questions. However, the authors
   over-claim one of their major results (see #7, below), along with saying some
   of their results are "significant" without providing statistical measure. (TODO: Pedro, mas acho que se pensarmos juntos mitigaremos)

### Reviewer #9C

#### Explicit Questions

1. Why do you call it a Likert-like scale? (TODO: Paulo)
2. How were the APIs introduced in the course? (TODO: Alfredo)
3. What were the exercises? (TODO: Pedro, mas do item 1 do revisor 2)
4. Is it possible that differences in recognition depend on the presentation of
   the APIs or teaching material, etc.? (TODO: Raphael -- pode-se colocar nas limitações do trabalho, relacionado ao item 5 do revisor 1)

@@

#### Other Comments

1. I personally think that a much larger discussion section is required that
  actually discusses the chosen method and the conclusions drawn from it. (TODO: Pedro)
2. Since you have a Likert-Scale, you can apply non-parametric tests on it in
   order to check, whether group A or B is different. Descriptive data and raw
   measurement is fine, because it gives readers a change to get an idea about
   the data, but only relying on such things is not enough. Having said this, I
   do not think that you can conclude something from the present paper, because
   the conclusion is just based on "looking at descriptive data". (TODO: Paulo)
3. Please fix the statistical part of the paper. Nothing would be more annoying
   than having a paper published whose interpretation depends on results that
   would be different in case appropriate analysis techniques would have been
   applied. (TODO: Pedro, mas acho que se pensarmos juntos mitigaremos)

### 2017-10-11 sections/02-apis.tex
Listing to Figure

The \textit{Pthreads} library controls thread execution by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin a parallel
region.  Listing~\ref{lst:pthreads} presents a \textit{Pthreads}
example. The code shows the parallel computation of the dot product of two
arrays.

The program in Listing~\ref{lst:pthreads} runs in a single thread until
line 22, when \texttt{pthread\_create} is called.  Each call
in the loop initializes a new thread.
The parallelization in this example divides the arrays in chunks
and shares multiplications between threads.
One thread synchronization tool available in \textit{Pthreads} is the barrier,
implemented as calls to \texttt{pthread\_join}. A thread that reaches a barrier
waits for other threads. Line 27 shows an example of thread synchronization
using a barrier.
To avoid race conditions on the \texttt{dot} shared variable
the example uses \textit{mutexes} to create a critical
region, as seen in lines 42 and 44.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{Pthreads} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.5cm,
                   label=lst:pthreads, keywords={%
                       DATATYPE, pthread_t, pthread_create,
                       pthread_join, task_function, NULL, int, main,
                       void, printf, return, pthread_mutex_t,
                       pthread_attr_t, pthread_attr_init,
                       MAX_THREADS, SIZE, char, struct, malloc,
                       MIN, pthread_mutex_lock, pthread_mutex_unlock,
                       pthread_exit%
                       },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                   ]
int A[SIZE], B[SIZE]; int dot = 0;
pthread_mutex_t lock;

@@

  /* Critical region */
  pthread_mutex_lock(&lock);
  dot += partial_dot;
  pthread_mutex_unlock(&lock);
  pthread_exit(0);
}
\end{lstlisting}

@@

Listing \ref{lst:omp} presents an \textit{OpenMP} example for the
same computation shown in Listing \ref{lst:pthreads}. The
declaration of a parallel region with the \texttt{firstprivate} pragma on line
1 indicates the variable \texttt{aux\_dot} should be private to every thread
and initialized with the enclosing scope's value.  Line 3 will be executed by
only one thread, because of the \texttt{single} pragma.
The loop in line 8 will be executed in parallel, with automatic work
division.  The pragma at line 12 defines a critical region which can only be
executed by a single thread at a time. The pragma at line 15 defines a
region that should only be executed by the master thread.

Note that the examples shown in Listings \ref{lst:pthreads} and
\ref{lst:omp} solve the same problem. The example in \textit{Phthreads}
contains much more code than the example in \textit{OpenMP}, and seems
to be much more complex as well. It is therefore understandable that students might
feel overwhelmed by \textit{Pthreads} example, especially when both examples are
presented side-by-side.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{OpenMP} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.6cm,
                   label=lst:omp, keywords={%
                       \#pragma,
                       omp, parallel, firstprivate,
                       single, omp_get_num_threads,
                       for, schedule, auto,
                       critical, master,
                       NULL, int, main,
                       void, printf, return%
                   },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                       ]
#pragma omp parallel firstprivate(aux_dot){
  #pragma omp single
  printf("Start of parallel region,
          number of threads: %d\n",
          omp_get_num_threads ());

@@

  #pragma omp master
  printf("Result: %d.\n", dot);
}
\end{lstlisting}

>>>>

The \textit{Pthreads} library controls thread execution by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin a parallel
region.  Figure~\ref{lst:pthreads} presents a \textit{Pthreads}
example. The code shows the parallel computation of the dot product of two
arrays.

The program in Figure~\ref{lst:pthreads} runs in a single thread until
line 22, when \texttt{pthread\_create} is called.  Each call
in the loop initializes a new thread.
The parallelization in this example divides the arrays in chunks
and shares multiplications between threads.
One thread synchronization tool available in \textit{Pthreads} is the barrier,
implemented as calls to \texttt{pthread\_join}. A thread that reaches a barrier
waits for other threads. Line 27 shows an example of thread synchronization
using a barrier.
To avoid race conditions on the \texttt{dot} shared variable
the example uses \textit{mutexes} to create a critical
region, as seen in lines 42 and 44.

\begin{figure}[htb]
\begin{minipage}{\linewidth}
\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   numberstyle=\tiny,
                   xleftmargin=0.5cm, frame=trBL,
                   keywords={%
                       DATATYPE, pthread_t, pthread_create,
                       pthread_join, task_function, NULL, int, main,
                       void, printf, return, pthread_mutex_t,
                       pthread_attr_t, pthread_attr_init,
                       MAX_THREADS, SIZE, char, struct, malloc,
                       MIN, pthread_mutex_lock, pthread_mutex_unlock,
                       pthread_exit%
                       },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                   ]
int A[SIZE], B[SIZE]; int dot = 0;
pthread_mutex_t lock;

@@

  /* Critical region */
  pthread_mutex_lock(&lock);
  dot += partial_dot;
  pthread_mutex_unlock(&lock);
  pthread_exit(0);
}
\end{lstlisting}
\end{minipage}
\caption{\textit{Pthreads} Sample Program}
\label{lst:pthreads}
\end{figure}

@@

Figure \ref{lst:omp} presents an \textit{OpenMP} example for the
same computation shown in Figure \ref{lst:pthreads}. The
declaration of a parallel region with the \texttt{firstprivate} pragma on line
1 indicates the variable \texttt{aux\_dot} should be private to every thread
and initialized with the enclosing scope's value.  Line 3 will be executed by
only one thread, because of the \texttt{single} pragma.
The loop in line 8 will be executed in parallel, with automatic work
division.  The pragma at line 12 defines a critical region which can only be
executed by a single thread at a time. The pragma at line 15 defines a
region that should only be executed by the master thread.

Note that the examples shown in Figures \ref{lst:pthreads} and
\ref{lst:omp} solve the same problem. The example in \textit{Phthreads}
contains much more code than the example in \textit{OpenMP}, and seems
to be much more complex as well. It is therefore understandable that students might
feel overwhelmed by \textit{Pthreads} example, especially when both examples are
presented side-by-side.

\begin{figure}[htb]
\begin{minipage}{\linewidth}
\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   numberstyle=\tiny,
                   xleftmargin=0.6cm, frame=trBL,
                   keywords={%
                       \#pragma,
                       omp, parallel, firstprivate,
                       single, omp_get_num_threads,
                       for, schedule, auto,
                       critical, master,
                       NULL, int, main,
                       void, printf, return%
                   },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                       ]
#pragma omp parallel firstprivate(aux_dot){
  #pragma omp single
  printf("Start of parallel region,
          number of threads: %d\n",
          omp_get_num_threads ());

@@

  #pragma omp master
  printf("Result: %d.\n", dot);
}
\end{lstlisting}
\end{minipage}
\caption{\textit{OpenMP} Sample Program}
\label{lst:omp}
\end{figure}
